---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

Todo: background and target for this document

## Loading and preprocessing the data

Dplyr package will be used for efficient data frame modifications.  
Analysis uses the activity monitoring dataset.

```{r part1}
    library(dplyr)
    data <- read.csv("activity.csv")
    ## TODO: Data processing
```

## What is mean total number of steps taken per day?

We start analysis by calculating mean and median of total steps per day. We also display histogram of total steps per day.

Following code uses loaded activity dataset, calculates total steps per day and displays histogram.

```{r totalSteps}
    
    ## Create table showing steps per day and calculate total
    date_data = group_by(data, date)
    steps_per_day <- summarise(date_data, total = sum(steps))
    
    ## print histogram showing steps per day
    hist (steps_per_day$total, main="Histogram of total steps per day", xlab = "Total steps per day")

```

Next we calculate mean and median for total steps per day.
```{r meanAndMedian}
    ## Calculate mean and median
    mean <- mean(steps_per_day$total, na.rm = TRUE)
    sprintf("Mean steps per day: %s", mean)
    ## Calculate median
    median <- median(steps_per_day$total, na.rm = TRUE)
    sprintf("Median steps per day: %s", median)
    
```


## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
